---
title: "two_step_factor_analysis"
author: "Lifan Liang"
date: "2022-08-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Factor analysis was performed with the unguided GSFA script Yifan provided. The setting of beta (the effects of guideRNAs on latent factors) is shown in the table below. 1 guide affects 3 factors, while other guides only affect one factor each. The 7th row is the intercept, which is set to 0 in this experiment.

```{r, echo=F}
beta.seurat <- read.table("data/Beta_unguided.txt")
rownames(beta.seurat) <- c(paste0("guide", 1:6), "Intercept")
colnames(beta.seurat) <- paste0("factor", 1:10)
knitr::kable(beta.seurat)
```

DEGs for a factor were identified with PIP ($Pr(F=1)$) > 0.95.

Factors were associated to guideRNAs via linear regression. Each factor values (posterior mean of Z) was regressed against the guides. P values of the coefficients were calculated. FDR was calculated across all the 60 P values (6 guides by 10 factors) with Benjamini-Hochberg method. An association between a guideRNA and a factor was statistically significant if FDR<0.05.

The set of DEGs for a guideRNA is the union of DEGs for its associated latent factors.

### Guided vs. unguided in Normal based simulation data

I compared the unguided with the guided GSFA on 300 simulation datasets (Normal based) with factor density 0.05. Guided GSFA has lower FPR and higher FNR. Unguided GSFA has increased false positives when a guide affects multiple factors.

#### Normal based fales positive rates and false negative rates
```{r, echo=F}
load("data/fit_normal_fpr_fnr.Rdata")
par(mfrow=c(1,2))
boxplot(c(fpr.guided), c(fpr.unguided), names=c("guided","unguided"), main="False Positive Rate")
boxplot(c(fnr.guided), c(fnr.unguided), names=c("guided","unguided"), main="False Negative Rate")
```

#### Normal based empirical false discovery rates and power
```{r, echo=F}
load("data/fit_normal_precision_recall.Rdata")
par(mfrow=c(1,2))
boxplot(c(1-precision.guided), c(1-precision.unguided), names=c("guided","unguided"), main="False Discovery Rate")
boxplot(c(recall.guided), c(recall.unguided), names=c("guided","unguided"), main="Power")
```

When we evaluated the false discovery rates across the 300 simulation datasets for each guideRNAs, the two-step approach has many more false discoveries than GSFA.

#### Number of simulation datasets where over 5% of detected DEGs were false.
```{r,echo=F}
library(ggplot2)
fdr.05 <- data.frame(method=c(rep("guided",6),rep("unguided",6)), guideRNA=rep(paste0("guide",1:6),2),FDR.05=c(colSums(precision.guided<0.95, na.rm=T), colSums(precision.unguided<0.95, na.rm=T)))
ggplot(fdr.05, aes(factor(guideRNA), FDR.05, fill=method)) + geom_bar(stat="identity", position = "dodge") + scale_fill_brewer(palette = "Set1") + ylab("# of datasets with FDR>5%")
```

#### Number of simulation datasets where <80% of true DEGs were detected.
```{r,echo=F}
library(ggplot2)
fdr.05 <- data.frame(method=c(rep("guided",6),rep("unguided",6)), guideRNA=rep(paste0("guide",1:6),2),power.8=c(colSums(recall.guided<0.8, na.rm=T), colSums(recall.unguided<0.8, na.rm=T)))
ggplot(fdr.05, aes(factor(guideRNA), power.8, fill=method)) + geom_bar(stat="identity", position = "dodge") + scale_fill_brewer(palette = "Set1") + ylab("# of datasets with power<80%")
```


### Guided vs. unguided in count based simulation data

In the count based scenario, results were similar. Both approaches perform worse than Normal base datasets. Two-step factor analysis has more outliers with much higher false positive rates. GSFA has more outliers with higher false negative rate. Unguided GSFA has increased false negatives when a guide affects multiple factors (e.g. guide 6 in the simulation).

#### Count based false positive rates and false negative rates.
```{r, echo=F}
load("data/fit_count_fpr_fnr.Rdata")
par(mfrow=c(1,2))
boxplot(c(fpr.guided), c(fpr.unguided), names=c("guided","unguided"), main="False Positive Rate")
boxplot(c(fnr.guided), c(fnr.unguided), names=c("guided","unguided"), main="False Negative Rate")
```

#### Count based empirical false discovery rates and power
```{r, echo=F}
load("data/fit_count_precision_recall.Rdata")
par(mfrow=c(1,2))
boxplot(c(1-precision.guided[,6]), c(1-precision.unguided[,6]), names=c("guided","unguided"), main="False Discovery Rate")
boxplot(c(recall.guided[,6]), c(recall.unguided[,6]), names=c("guided","unguided"), main="Power")
```

#### Number of simulation datasets where over 5% of detected DEGs were false.
```{r,echo=F}
library(ggplot2)
fdr.05 <- data.frame(method=c(rep("guided",6),rep("unguided",6)), guideRNA=rep(paste0("guide",1:6),2),FDR.05=c(colSums(precision.guided<0.95, na.rm=T), colSums(precision.unguided<0.95, na.rm=T)))
ggplot(fdr.05, aes(factor(guideRNA), FDR.05, fill=method)) + geom_bar(stat="identity", position = "dodge") + scale_fill_brewer(palette = "Set1") + ylab("# of datasets with FDR>5%")
```

#### Number of simulation datasets where <80% of true DEGs were detected.
```{r,echo=F}
library(ggplot2)
fdr.05 <- data.frame(method=c(rep("guided",6),rep("unguided",6)), guideRNA=rep(paste0("guide",1:6),2),power.8=c(colSums(recall.guided<0.8, na.rm=T), colSums(recall.unguided<0.8, na.rm=T)))
ggplot(fdr.05, aes(factor(guideRNA), power.8, fill=method)) + geom_bar(stat="identity", position = "dodge") + scale_fill_brewer(palette = "Set1") + ylab("# of datasets with power<80%")
```